Anomaly detection with kernel density estimation on manifolds

### 2022-05-22 paper/paper.tex
Format

The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\). This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.

The rest of the paper is organized as follows.
In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
Conclusions and discussions are presented in \autoref{conclusion}.
Readers interested in the the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.

@@

In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation}
where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\). When using the standard gaussian kernel \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).

@@

  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;

  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;

  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};

  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.

@@

We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means \(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\). The mixture proportions are equally set as \(\pi_i=0.25, i=1,2,3,4\).
Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
\begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation}
\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities. We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow. Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels. These are \emph{true outliers} to be compared with outliers from the kernel density estimates.

@@

Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg. 1\%, 5\%, 50\%, 99\%, \textgreater99\%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix. Matrix \(\pmb{Q}\) satisfies
\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\). Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms. Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).

@@

The tangent space \(T_pM\) also admits a dual space \(T^*_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector} \(z_p: T^*_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).

@@

The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\).
\(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
For a curve \(\gamma: I \rightarrow M\), the length of the curve is
\[l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt}\],
where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
\[
Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
\]
which is also called the \emph{Riemannian measure} on \(M\).

@@

Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors.
Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\). The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
\[
<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
\]
where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\).
The tangent vectors \(f_*u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\).
Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).

@@

\hypertarget{volumn-density-function}{%
\subsection{Volumn density function}\label{volumn-density-function}}

@@

\hypertarget{example-s1}{%
\paragraph{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}

@@

We use the density estimator in \textcite{Pelletier2005-vu}.
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
\]
where \(d=2\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

@@

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neightborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
\]
The bandwidth \(h\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.

>>>>

The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold. In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(M\). This kernel density estimate can then be used to produce the highest density region plots\autocite{Hyndman1996-lk} for outlier visualization.

The rest of the paper is organized as follows. In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth. \autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.

@@

In general, given a multivariate random sample \(\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d\), \(i = 1, 2, \dots, N\) drawn from a density \(f\), the variable kernel density estimate of \(f\) at point \(\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d\) is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation}
where \(\pmb{H}_i\) is called the bandwidth matrix which is a \(d\times d\) symmetric positive definite matrix variable at each \(i\), \(K(\pmb{y})\) is the kernel function, and \(K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})\).
When using the standard gaussian kernel, \(K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})\), we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation}
with \(|\pmb{H}_i|\) being the determinant of matrix \(\pmb{H}_i\).

@@

  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \\ref{equ:denestimator};
  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
  \end{algorithmic}
\end{algorithm}

@@

We first generate a 2-dimensional data of size \(N=2000\) from a Gaussian mixture model with four components with different means
\(\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime\) and the same variance-covariance matrix \(\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4\). The mixture proportions
are equally set as \(\pi_i=0.25, i=1,2,3,4\).
Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as
\begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation}

\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \eqref{eq:gmm}, with brighter colors showing higher densities and darker colors showing lower densities. We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow. Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels. These are \emph{true outliers} to be compared with outliers from the
kernel density estimates.

@@

Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg. 1\%, 5\%, 50\%, 99\%, \textgreater99\%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue. From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all
manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\)(denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms. The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\). For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix. Matrix \(\pmb{Q}\) satisfies
\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the pairwise Euclidean distances between data points in
\(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\).
Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms. Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).

@@

The tangent space \(T_pM\) also admits a dual space \(T^\star_pM\) called the \emph{cotangent space} with corresponding \emph{cotangent vector}
\(z_p: T^\star_pM \rightarrow \mathbb{R}^d\), and basis denoted as \(dx_1(p),\dots,dx_d(p)\).

@@

The Riemannian metric \(g\) can be used to define the norm of a vector \(u\), \(\|u\| = \sqrt{<u,v>_g}\) and the angle between two vectors \(u\) and \(v\), \(\cos\theta = \frac{<u,v>_g}{\|u\| \|v\|}\), which are the geometric quantities induced by \(g\). \(g\) could also be used to define the line element \(dl^2 = g_{ij}dx_idx_j\) and the volume element \(dV_g = \sqrt{\det(g)}dx_1\dots dx_d\), where \((x_1,\dots,x_d)\) are the local coordinates of the chart \((U, \varphi)\).
For a curve \(\gamma: I \rightarrow M\), the length of the curve is
\[
l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx_i}{dt} \frac{dx_j}{dt} dt},
\]
where \(\gamma(I) \subset U\). The volume of \(W \subset U\) is defined as
\[
Vol(W) = \int_W \sqrt{\det(g)}dx_1 \dots dx_d,
\]
which is also called the \emph{Riemannian measure} on \(M\).

@@

Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. Let \(f\) be an embedding from the Riemannian manifold \((M,g)\) to another smooth manifold \(E\). The pushforward \(h=\varphi*g\) of the Riemannian metric \(g\) along \(\varphi \equiv f^{-1}\) is a linear map \(f_*:TM \rightarrow TE\), which maps the tangent vectors \(u_p\) at point \(p \in M\) to the tangent vectors \(f_*u_p\) at the mapping point \(f(p) \in E\), and satisfies that for \(u, v \in T_f(p)E\),
\[
<u,v>_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
\]
where \(df_p^{-1}\) is the Jacobian of \(f^{-1}\). The tangent vectors \(f_\star u_p\) are equivalent to the velocity vector of a curve \(\gamma: I\rightarrow M\) passing through point \(p\) at time zero with a constant speed \(\gamma^{-1}(0)=u_p\). Similarly, the pullback maps the cotangent vectors \(z_{f(p)}\) at \(f(p) \in E\) to cotangent vectors at \(p \in M\) acting on tangent vectors \(u \in T_pM\).

@@

\hypertarget{volume-density-function}{%
\subsection{Volume density function}\label{volume-density-function}}

@@

\hypertarget{example-s1}{%
\subsection{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}

@@

We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
\]
where \(d=1\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

@@

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neighborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
\]
where the bandwidth \(r\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.

\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}

Example of \(S^2\) in \textcite{Henry2009-om}.

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
\[
\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
\]
if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we take the partial derivatives of \$\textbackslash Psi\^{}\{-1\}(s,t)\$ with respect to \$s\$ and \$t\$ and have that
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
\]
\[
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
\]
Then each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain that the volume density on the sphere is
\[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\]
for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

Similarly in \(S^1\), we could let \(\{v, p\}\) be the basis of \(\mathbb{R}^2\). Then the exponential chart induced

Polar coordinates:
\(\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v\).

Cartesian coordinates:
\[
\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
= - \sqrt{1 - p^2}v
\]
Then the volume density function is
\[
\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
\]
and the geodesic distance is
\[
d_g(p,q) = \arccos(p) - \arccos(q).
\]
